\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-11 doc/publications/plos16/plos16.tex
changes to plos16.tex

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in our RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.

@@

So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
enable these increasingly large vectors to be loaded into memory.
\marginpar{not clear. Are RDDs the problem or the solutuion Aidan: Better?}

On the other hand, \cursedforest\ is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node
of a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each
of these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the
results are reduced to return the feature which gives the greatest information gain.  This process is then repeated
until \cursedforest\ has created the entire decision tree.

The current implementation of \cursedforest\ uses an ``Gini impurity''
criteria for splitting. Let $f_q$ be the fraction of items labelled
with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

>>>>

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.

@@

So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
enable these increasingly large vectors to be loaded into memory.

On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
entire decision tree.

The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}
